---
title: "Coursera  Reproducible Research - Project 1"
author: "Chandrakanth Saravanan"
date: "December 29, 2017"
output: html_document
---

##Read Data
Code to read data


```{r Project 1, echo = TRUE}
activity <- read.csv("activity.csv", header = TRUE)

```

## Q1) What is the total number of steps taken per day?

Code to calcualte total number of steps taken per day
```{r steps per day, echo=TRUE}
StepsPerDay <- tapply(activity$steps, activity$date, sum)
StepsPerDay
```

## Make a histogram of the total number of steps taken each day
```{r, echo=TRUE}
hist(StepsPerDay, xlab = "Number of Steps", main = "Histogram: Steps per Day")

```

## Calcualte and report the mean and median of the total number of steps taken per day
```{r , echo=TRUE}
MeanPerDay <- mean(StepsPerDay, na.rm = TRUE)
MedianPerDay <- median(StepsPerDay, na.rm = TRUE)
```
The mean steps per day is 10766.186 steps and the median number of steps per day is 10765 steps.

## Q2) What is the average daily acitivity pattern?
```{r , echo=TRUE}
StepsPerInterval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
plot(as.numeric(names(StepsPerInterval)), 
     StepsPerInterval, 
     xlab = "Interval", 
     ylab = "Steps", 
     main = "Average Daily Activity Pattern", 
     type = "l")
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxInterval <- names(sort(StepsPerInterval, decreasing = TRUE)[1])
maxSteps <- sort(StepsPerInterval, decreasing = TRUE)[1]

```

The interval with maximum activity is 83, at 206.169 steps
## Q3) Impute missing values

Calcualte and report the total number of missing values in the dataset
```{r, echo=TRUE}
NA.vals <- sum(is.na(activity$steps))
NA.vals
```

## Devise a stratergy for filling in all of the missing values in the dataset

```{r, echo=TRUE}
StepsPerInterval <- tapply(activity$steps, activity$interval, mean, na.rm = TRUE)
# split activity data by interval
activity.split <- split(activity, activity$interval)
# fill in missing data for each interval
for(i in 1:length(activity.split)){
    activity.split[[i]]$steps[is.na(activity.split[[i]]$steps)] <- StepsPerInterval[i]
}
activity.imputed <- do.call("rbind", activity.split)
activity.imputed <- activity.imputed[order(activity.imputed$date) ,]
```

## Make a histogram of the total number of steps taken and calcualte and report the mean and median total number of steps taken per day

```{r, echo=TRUE}
StepsPerDay.imputed <- tapply(activity.imputed$steps, activity.imputed$date, sum)
hist(StepsPerDay.imputed, xlab = "Number of Steps", main = "Histogram: Steps per Day (Imputed data)")
```

```{r, echo = TRUE}
MeanPerDay.imputed <- mean(StepsPerDay.imputed, na.rm = TRUE)
MedianPerDay.imputed <- median(StepsPerDay.imputed, na.rm = TRUE)
MeanPerDay.imputed
MedianPerDay.imputed
```

## Q4) Arer there differences in acitivity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
activity.imputed$day <- ifelse(weekdays(as.Date(activity.imputed$date)) == "Saturday" | weekdays(as.Date(activity.imputed$date)) == "Sunday", "weekend", "weekday")
```

Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days.

```{r, echo=TRUE}
# Calculate average steps per interval for weekends
StepsPerInterval.weekend <- tapply(activity.imputed[activity.imputed$day == "weekend" ,]$steps, activity.imputed[activity.imputed$day == "weekend" ,]$interval, mean, na.rm = TRUE)

# Calculate average steps per interval for weekdays
StepsPerInterval.weekday <- tapply(activity.imputed[activity.imputed$day == "weekday" ,]$steps, activity.imputed[activity.imputed$day == "weekday" ,]$interval, mean, na.rm = TRUE)

# Set a 2 panel plot
par(mfrow=c(1,2))

# Plot weekday activity
plot(as.numeric(names(StepsPerInterval.weekday)), 
     StepsPerInterval.weekday, 
     xlab = "Interval", 
     ylab = "Steps", 
     main = "Activity Pattern (Weekdays)", 
     type = "l")

# Plot weekend activity
plot(as.numeric(names(StepsPerInterval.weekend)), 
     StepsPerInterval.weekend, 
     xlab = "Interval", 
     ylab = "Steps", 
     main = "Activity Pattern (Weekends)", 
     type = "l")
```

